\chapter{Curve Selection}

Let $E$ be an elliptic curve defined over a finite field $K$.
Let $P \in E(K)$ be a point of prime order $r$. Let $G = \langle P \rangle$.

If $L$ is a field extension of $K$
that contains the $r$th roots of unity (so $r$ divides $|L| - 1$),
then the Tate pairing
on $G$ can be computed by performing operations in $L$.

On the other hand,
let $E[r]$ denote the group of points of order $r$.
The Weil pairing is defined on $E[r]$. For some field extension
$L'$ of $K$, $E(L')$ contains $E[r]$, and the Weil pairing can be
computed by performing operations in $L'$.

In order for the Tate pairing to be efficiently computable,
operations must be efficient in $L$.
Similarly, for the Weil pairing to be efficiently
computable, operations must be efficient in $L'$.

Thus we seek fields $L$ or $L'$ that are small enough so that field
operations are still fast.

It turns out that such a field $L'$ will always contain the $n$th
roots of unity, but the converse is not true: it is possible
for a field $L$ to contain the $r$th roots of unity but
at the same time part of $E[r]$ lies outside $E(L)$.

However, the converse is almost true:

\section {The Embedding Degree}

\begin{theorem} \cite{BK}
Let $E$ be an elliptic curve defined over $\mathbb{F}_q$.
Let $r$ be a prime dividing $N = \#(\mathbb{F}_q$) with
$r \nmid q-1$. Then $E(\mathbb{F}_{q^k})$ contains all $r^2$
points of order $r$ if and only if $r \mid q^k - 1$.
\end{theorem}

\begin{proof}
It is well-known that if $E(\mathbb{F}_{q^k})$ contains $E[r]$
then $r \mid q^k - 1$, even without assuming $r \mid N$ or
$r \nmid q-1$.

Conversely, suppose $k > 1$ and $r \mid q^k - 1$.
Let $\Phi$ denote the Frobenius map. Consider the subgroup $T$ of $E[r]$
consisting of all points of trace zero, that is
\[
T = \{ Q \in E[r] : \tr Q = Q + \Phi(Q) + ... + \Phi^{k-1}(Q) = O \}
\]
(The group $T$ may be explicitly constructed by using the map $P \mapsto
P - \Phi(P)$ on points of $E(\mathbb{F}_{q^k})$.)
Now we have $\Phi(T) = T$, and also $T$ is not contained
in $E(\mathbb{F}_q)$ by assumption.

Hence $T$ is an eigenspace of $\Phi$, but not the $1$-eigenspace. Since the
eigenvalues of $\Phi$ are $1$ and $q$, we see that $T$ must be the
$q$-eigenspace of $\Phi$ and hence
\[
\Phi^k(Q) = q^k Q = Q
\]
since $r | q^k - 1$. Thus $T$, like $E(\mathbb{F}_q)$ is fixed under
$\Phi^k$, and since these groups are linearly independent they generate all
of $E[r]$, implying that all of $E[r]$ is fixed under $\Phi^k$. Hence
$E[r] \subset E(\mathbb{F}_{q^k})$.
\end{proof}

The $k$ in the above theorem is referred to as the \emph{embedding degree}.

\begin{definition}
Let $E$ be an elliptic curve defined over $\mathbb{F}_q$.
Let $P \in E(K)$ be a point of prime order $r$.
Let $G = \langle P \rangle$.
The \emph{embedding degree} of $G$ is the smallest positive integer $k$
such that $r \mid q^k - 1$.
\end{definition}

Thus we have shown that if the embedding degree $k$ is greater than 1,
then both the Tate pairing and the Weil pairing may be computed by
performing field arithmetic in $\mathbb{F}_{q^k}$.

However, if the embedding degree $k$ is 1, the Tate pairing may
be computable in $\mathbb{F}_q$, but the Weil pairing may not be.
In practice this is not significant because we shall see we prefer
$k > 1$. However, the Tate pairing turns out to have other benefits.

For example, consider the curve $E: Y^2 = X^3 + X + 6$ which has
18 points over $\mathbb{F}_{19}$.
The point $R=(0,5)$ generates a cyclic group of order 18,
so the point $P = 6R = (12,13)$ generates a cyclic group of order 3.

We see $E[3]$ cannot be contained in $\langle R \rangle = E(\mathbb{F}_{19})$
since $E[3]$ is not cyclic, thus the Weil pairing cannot be computed.
(It turns out we must move to $\mathbb{F}_{19^3}$ to do so.)

In contrast, $3 \mid 19-1$ hence the Tate pairing can be
computed in $\mathbb{F}_{19}$. In fact, we find $e(P, P) = 7$, where
$e$ denotes the Tate pairing.

\section { Security Considerations }

Thus our goal is to find curves $E(\mathbb{F}_q)$ containing a subgroup
$G$ of prime order $r$ with embedding degree $k$ small enough so that
operations in $\mathbb{F}_{q^k}$ are efficiently computable.
All other things being equal, we prefer $q$ to be small as possible so
that operations in $G$ are fast as possible and can be represented by
as few bits as possible.

On such curves, the Tate pairing is always efficiently computable.
If $k > 1$ then the Weil pairing is also efficiently computable.

On the other hand, we must consider the lower bounds on $q$ and $q^k$.
Recall elliptic curves are attractive for cryptography
because unlike finite fields,
no specialized discrete logarithm algorithm is known for a general elliptic
curve.

Generally speaking, for a cryptographically useful elliptic curve
we need only ensure that $r$ is large enough to foil generic attacks
(e.g. Pollard rho, baby-step-giant-step).

However, a low embedding degree $k$ complicates matters. A bilinear map
sends points of an elliptic curve to elements of a finite field
$\mathbb{F}_{q^k}$. In particular, given a subgroup $G$ of $E(\mathbb{F}_q)$
of prime order $r$, by bilinearity and nondegeneracy we have
\[ \dlog (P, nP) = \dlog ( e(P,P), e(P, nP) )
= \dlog (e(P,P),e(P,P)^n) = n \]
for $n \in \{0,...,r-1\}$.
Thus if the discrete logarithm problem can be solved
in $\mathbb{F}_{q^k}$, then it can also be solved in $G$ \cite{mov, fr}.

Hence we must also ensure $q^k$ is large enough so that
finite field discrete logarithm algorithms such as index calculus are
infeasible in $\mathbb{F}_{q^k}$.

The curves used in pairing-based cryptography are special in other senses.
They are supersingular, or have complex multiplication. This possibly means
they are vulnerable, but no specific attacks for either case are known.

We have now described all necessary conditions on the sizes of
$r$ and $q^k$:

\begin{enumerate}
\item
$r$ must be a large enough prime so that generic discrete logarithm attacks
in a group of order $r$ are ineffective. Since $q \approx \#E(\mathbb{F}_q)$,
this places a similar lower bound on $q$.
\item
$q$ ought to be as small as possible, so that computations in $\mathbb{F}_q$
are as fast as possible.
\item
$q^k$ must be large enough so that finite field discrete logarithm attacks
in $\mathbb{F}_{q^k}$ are ineffective.
\item
$q^k$ must be small enough so that operations in $\mathbb{F}_{q^k}$
are efficient. $q^k$ should be small as possible so that operations
are as fast as possible.
\end{enumerate}

Of course, the first three statements are true for any cryptographically
useful elliptic curve, not just for pairing-based cryptography.

Currently it is considered acceptable to have $r$ about 160-bits.
As for $q^k$, 1024 bits is adequate for many applications, and calculations
in fields of this size can certainly be performed.
Ideally we have $r \approx \#E(\mathbb{F}_q) \approx q$
and hence $q$ is also about 160 bits. This gives an
embedding degree $k$ around $1024 / 160 = 6.4$.

We shall see how to construct curves of embedding degree 6 and 12.
For the near future, embedding degree 6 should be reasonable if used along
with an $r$ that is at lesat $170$ bits long,
and embedding degree 12 curves may be more desirable as time passes.

\section { Approaches to Finding Curves }

A randomly-chosen elliptic curve will have a large embedding
degree, giving resistance to the aforementioned discrete logarithm
attack, but also making it useless for pairing-based cryptography.
We must take one of two approaches to find curves suitable for pairings:

\begin{enumerate}
\item
Supersingular curves are guaranteed to have a small embedding degree,
and are easy to construct. They have been completely classified. Operations
on some of them can be highly optimized.
\item
By carefully tailoring the complex multiplication method of constructing
elliptic curves, we can produce curves of a certain embedding degree. 
\end{enumerate}

We note that supersingular hyperelliptic curves have also been considered,
and it is thought that using the hyperelliptic equivalent of the
complex multiplication method may lead to more curves with low
embedding degrees, though further research in this area is required.

\section { Supersingular Curves }

\section { Complex Multiplication }
