\chapter {Elliptic Curves}

We review basic facts about elliptic curves.
Aside from the problem of generating suitable elliptic curves
and counting the size of the resulting group,
we will cover enough theory to replace finite fields with
elliptic curves in cryptosystems based on cyclic groups.

We discuss pairings in the following chapter, and describe algorithms
to find pairing-friendly curves in the chapter after that. We will not
discuss curve-finding and point-counting algorithms that are geared
towards standard elliptic curve cryptography (which necessarily requires
curves that are not pairing-friendly), but instead direct
the interested reader to the references~\cite{bss}.

\section {Informal Overview of Elliptic Curves}

We present a highly informal overview to elliptic curve theory,
which may aid those that have not encountered them before.
This section can be safely skipped.

Consider a polynomial $C$ in two variables $X, Y$.
We are interested in the solutions to $C = 0$
which describe a curve on a two-dimensional plane.

We first observe that if $C'$ is another curve that is an affine transformation
of $C$, that is, if we can linearly transform (e.g. rotate, scale, shear)
and then translate $C$ to
obtain $C'$ then a correspondence exists between
the solutions to $C = 0$ and the solutions to $C' = 0$. Knowing
the solutions of one allows us to easily compute the solutions of the other,
For this reason we consider $C$ and $C'$ to be equivalent.

If every term in $C$ has combined degree of at most 1, that is,
if $C = a X + b Y + c$ then $C$ describes a line. The geometry of lines
is too simple to yield anything cryptographically useful.

If every term in $C$ has combined degree at most 2, then $C$ describes
a single line, a pair of lines, an ellipse, a parabola or a hyperbola.
The first two possiblities can be viewed as special cases, occuring
when $C$ is reducible or degenerate in some sense.

By adding \emph{points of infinity} to the plane, we can find affine
transformations that change any ellipse, parabola or hyperbola into the
unit circle centred at the origin. Intuitively, the two ends of the parabola
can be thought of as meeting
at a point at infinity, forming a circle,
and similarly opposite ends of hyperbolas connect at infinite points.

TODO: diagram?

Thus to study degree 2 curves is to study lines or circles, whose geometry
is again is too simple for our purposes.

However, degree 3 curves, called elliptic curves,
are quite nontrivial (for starters, unlike the previous two
cases cannot transform any (irreducible) elliptic curve into
any other)
and have a rich structure well-suited for cryptography.

For any irreducible elliptic curve $C$,
applying appropriate affine transformations
produces an equation $C'$ of a certain form known
as the \emph{Weierstrass form}.
In this form, $C'$ always contains exactly one point at infinity.
Transforming $C$ so
that no points are infinite is
possible but leads to more complicated
equations. Using an equation that has
exactly one point at infinity simplifies some of the resulting equations.

For example, the Fermat curve $X^3 + Y^3 = 1$ becomes
?!? after a suitable affine transformation.

\section {Points on Elliptic Curves}

Let $\Fq$ be a field for some prime $q > 3$.
Unless otherwise specified we shall always
define curves over a field of prime order and of characteristic greater
than three.
Elliptic curves can be implemented over fields of characteristic 2 and 3
and enjoy many optimizations,
but suffer from specialized discrete log attacks and should generally
be avoided.

An elliptic curve $E$ over such a field $\Fq$ is an equation of the form
\[ E: Y^2 = X^3 + a X + b \]
where $a, b \in \Fq$.
Let $\Delta = 4 a^3 + 27b^2$ (the discriminant of the cubic in $x$). Then
$E$ is \emph{singular} if $\Delta = 0$ (the cubic has a repeated root)
and \emph{nonsingular} otherwise (the cubic has distinct roots).

Unless otherwise stated we always consider nonsingular elliptic curves.
(Later we will encounter \emph{supersingular} curves, a particular
flavour of nonsingular curves and are \emph{not} singular despite their
name.
The antonym of supersingular is \emph{ordinary} or \emph{nonsupersingular}.)

For any field $\Fqk$ define $E(\Fqk)$ (or $E/\Fqk$)
to be the set of all solutions
of $E$ over $\Fqk$, called the \emph{finite points}
along with a special point denoted $O$,
that is called the \emph{point at infinity}. We write $\#E(\Fqk)$ 
or $|E(\Fqk)|$ for
the number of elements of $E(\Fqk)$.

Although irrelevant, the reader may be curious about the point at infinity.
Briefly, by considering the equation in
projective coordinates one can show that $O = (0,1,0)$ is always a unique
infinite solution to the equation. Intuitively,
the point $O$ can be thought of as the point where all lines
parallel to the Y-axis meet.

We quote two well-known theorems from the theory of elliptic curves.

\begin{theorem}
[Hasse] Let $t = q^k + 1 - \#E(\Fqk)$.
Then $|t| \le 2\sqrt{q^k}$.
\end{theorem}

Thus the number of points on an elliptic curve in a given field
is on the same order as the size of the field.
The quantity $t$ is called the \emph{trace of Frobenius}.

\begin{theorem}
[Weil] Let $t = q + 1 - \#E(\Fq)$
where $q$ is a prime power.
Factor the polynomial $x^2 - t x + q$ as $(x-\alpha)(x-\beta)$ over
$\mathbb{C}[x]$.
Then
\[ \#E(\Fqk) = q^k + 1 - (\alpha^k + \beta^k) .\]
\end{theorem}

This last theorem is more practical in the following form. Let $t_0 = 2$,
Let $t_1 = q + 1 - \#E(\Fq)$.
Define $t_n$ recursively by
\[ t_n = t_1 t_{n-1} - q t_{n-2} .\]
Then $\#E(\Fqk) = q^k + 1 - t_k$.

For example, consider the curve E\cite{bk} given by
\[ Y^2 = X^3 + X + 6 \]
over $\F_{19}$. There are 18 points
\[
\begin{array}{llllll}
(0,5), & (4,6), & (2,4), & (3,6), & (14,3), & (12,13), \\
(18,2), & (10,3), & (6,0), & (10,16), & (18, 17), & (12,6), \\
(14,16), & (3,13), & (2,15), & (4,13), & (0,14), & O
\end{array}
\]
thus the trace of Frobenius $t = 2$.

Over $\F_{{19}^2}$, we have
\[
\#E(\F_{{19}^2}) = 19^2 + 1 - t_2
\]
where $t_2 = 2\cdot 2 - 19 \cdot 2 = -34$, thus
$\#E(\F_{{19}^2}) = 396$.

\section {The Chord-Tangent Law of Composition}

TODO: pictures

We define an operation $+$ on $E(\Fqk)$.
Let $P = (a,b), Q = (c,d) \in E(\Fqk)$ be finite points.

If $P \ne Q$, then it is not hard to show that if $a \ne c$
then the line through $P$ and $Q$ must intersect $E$ at another point
$(x,y)$ where $x, y\in \Fqk$. Note that $(x,-y)$ also is a solution of $E$.
Define $P + Q = (x, -y)$ for $a \ne c$.
If $a = c$ (in which case we must have $b = -d$),
then define $P + Q = O$.

Now suppose $Q = P$. In this case, consider the tangent line going through
$P$. It turns out it must intersect $E$ at another point $(x,y)$ where
$x,y\in\Fqk$ unless $b = 0$. Define $P + P = (x, -y)$ for $b \ne 0$.
For $b = 0$ define $P + P = O$.

Lastly define $P + O = P$, $O + O = O$.

This operation turns $E(\Fqk)$ into a group.
The point $O$ is the identity, and the inverse
of a point $P = (x,y)$ is the point $-P = (x,-y)$.

As usual, for an integer $n$ define $0 P = O$, $1 P = P$,
$n P = (n-1)P + P$ for $n > 1$ and $n P = -(-n)P$ for $n < 0$.

Recall the previous chapter used mulitplicative group notation to emphasize
the connection
between discrete log cryptosystems and pairing-based cryptosystems:
we are using elliptic curve groups where one would use
the multiplicative group of a finite field.

Mathematicians use additive group notation for the elliptic curve group
(since the group is Abelian), and we will adhere to this convention in this
section.

\section {Explicit Formulas}

Let $E: y^2 = x^3 + ax + b$ be an elliptic curve over $\Fqk$. Let
$P_1, P_2 \in E(\Fqk)$ and suppose
we wish to find $P_3 = P_1 + P_2$.
Note the case where at least one of $P_1, P_2$ is $O$ is trivial,
so assume both $P_1$ and $P_2$ are finite.
Then write $P_1 = (x_1, y_1), P_2 = (x_2, y_2)$.

If $x_1 = x_2$ and $y_1 = -y_2$
then the line $V$ through $P_1$ and $P_2$ is vertical and
can be given by
\[ V : X - x_1 = 0  . \]
In this case $P_3 = O$ (and $P_1 = -P_2$).

Otherwise $P_3$ also must be finite, hence write $P_3 = (x_3, y_3)$.
We have two cases. If $P_1 = P_2$ then the tangent line $T$ at
$P_1$
has slope
\[\lambda =
\left[ \frac{\partial E / \partial X}{\partial E/ \partial Y} \right]_{(x_1,y_1)}
= \frac{3x_1^2 + a}{2y_1} . \]
and if $P_1 \ne P_2$ then the line $L$ through $P_1$ and $P_2$ has slope
\[ \lambda = (y_2 - y_1)/(x_2 - x_1) . \]

If we set $\mu = y_1 - \lambda x_1$ then
the equation of $L$ or $T$ is given by:

\[ L, T :  Y - (\lambda X + \mu) = 0 \]

We can find the $x$-coordinate of the other point of intersection $(x_3, -y_3)$
by substituting $Y = \lambda X + \mu$ into $E$. We find
\[ E(X, \lambda X+\mu) = X^3 - \lambda^2 X^2 + a_1 X + a_0 = 0 \]
for some $a_1, a_0 \in \Fqk$, thus the sum of the roots
$x_1 + x_2 + x_3 = \lambda^2$.

This allows us to compute $x_3, y_3$. Explicitly we have for $P_1 \ne P_2$:

\[
\begin{array}{rcl}
\lambda &\leftarrow& (y_2 - y_1)/(x_2 - x_1) \\
x_3 &\leftarrow& \lambda^2 - x_1 - x_2 \\
y_3 &\leftarrow& (x_1 - x_3) \lambda - y_1
\end{array}
\]
and for $P_1 = P_2$:
\[
\begin{array}{rcl}
\lambda &\leftarrow& (3x_1^2 + a)/(2y_1) \\
x_3 &\leftarrow& \lambda^2 - 2x_1 \\
y_3 &\leftarrow& (x_1 - x_3) \lambda - y_1
\end{array}
\]
The most expensive step is the division in the computation of
$\lambda$.

The expressions for the lines $L,T,V$ will be used later in the computation
of a pairing.

For the curve $Y^2 = X^3 + X + 6$ over $\F_{19}$,
using these formulas, it can be verified for instance that
$(0, 5) + (0, 5) = (4, 6)$,
and $(0, 5) + (2, 15) = (4, 13)$.

\section {Elliptic Curve Cryptography}

Having described an Abelian group operation on points on an elliptic curve,
we can almost see how elliptic curves may be used in cryptography.
Let $E$ be an elliptic curve over a field $K$.

Firstly, we can generate random points on $E$:
after generating a random $x \in K$, we can attempt to solve $E$ for $y$ by
using any finite field square root algorithm which we discuss below.
If no solution exists, simply pick a new random $x$.

Hashing to a point in $E$ is similar. The input is hashed to some
$x \in K$, and then a corresponding $y$ is sought. On failure,
a new $x$-coordinate is deterministically generated from $x$, and again
we attempt to solve $E$ for $y$. Repeating this process as many times as
necessary eventually yields a valid point $(x,y) \in E(K)$.

Let $F$ be a finite field.
The above formulas describe how to compute inverses and sums of
points. Thus it seems natural to replace elements of $F^*$
with elements of $E(K)$, and perform
point addition instead of modular multiplication.

The elliptic curve
equivalent of exponentiation is called
\emph{point multiplication}: $a P$ represents
adding $P$ to itself $a$ times just as $g^a$ represents $g$ multiplied
by itself $a$ times in finite fields.
Point multiplication can be performed efficiently via carefully
chosen point
additions, in a process that mirrors the repeated squaring technique
for exponentiation in finite fields.

Recall that cryptographic schemes in $F^*$ often operate
within a subgroup $G$ of a particular order $r$,
so elements chosen at random and hashed to must have
order $r$, or a factor of $r$. But an element of $F^*$
in general does not have the desired order.
Thus after choosing or hashing to some element $x \in F^*$,
to obtain an element of a suitable order one simply exponentiates $x$
by $n / r$ where $n = \#F^*$.

Similarly, the construction a point of order $r$ (or a factor of $r$)
from some given point $P \in E(K)$ is accomplished by multiplying $P$ by
$n / r$ where $n = \#E(K)$.

Assume for now that we know $n = \#E(K)$. Then from group theory (a
Sylow theorem), for any prime $r$ dividing $n$,
there exists a point $P \in E(K)$ of order $r$.

Then if $r^2$ does not divide $n$ then there is exactly one
subgroup $G$ of $E(K)$ of order $r$.

This suggests the following procedure for implementing any cryptographic
scheme based on cyclic groups of prime order:

\begin{enumerate}
\item
Choose any curve $E(K)$ and somehow work out $n = \#E(K)$.
\item
Find a prime $r$ dividing $n$, such that $r^2 \nmid n$.
We shall work in the unique cyclic subgroup $G \subset E(K)$ of points
of order $r$.
\item
When a random group element of $G$ is required, first choose a random point
of $E(K)$ and then multiply by $n / r$. Similarly, when hashing to
a point of $G$, first hash to a point in $E(K)$ and then mulitply by
$n / r$.
\item
Other operations are straightforward: every time a group
operation is required, we perform a point addition. To find an inverse
of a group element, we negate the $y$-coordinate of a point. When
an exponentiation is called for we carry out a point multiplication
\end{enumerate}

TODO: table? diagram?

In the running example, $E : Y^2 = X^3 + X + 6$ over $\F_{19}$,
the order of the group is $n = 18 = 3^2 \times 2$, and it so happens
that the group is cyclic: it can be checked that $(0,5)$ is a generator.
Thus this curve can be used
for cryptography on a cyclic group of order $r = 3$. However, in general,
if $r^2 | n$ there may be more than one subgroup of order $r$ and the
above procedure cannot be followed.

But how do we compute $\#E(K)$ for a given curve?
As mentioned earlier, fast algorithms exist for this~\cite{bss}
but it turns out that if a pairing is desired, we must seek out
elliptic curves whose orders satisfy various conditions.
As a result,
instead of choosing a curve first and counting the number of points $n$ and
hoping for a large prime factor $r$ of $n$,
we must use families of curves where the size of the group
is known in advance and has the requisite properties,
a subject of a later chapter.

For now, we exhibit one case where $\#E(K)$ is always easy to work out
and furthermore $E(K)$ is always cyclic.

\section {Singular Elliptic Curves}

Using singular elliptic curves
is equivalent to conventional cryptography.
It can be shown that the set $E_{ns}$ nonsingular points (all but
one) of a singular elliptic curve over a field $K$
also form a group under the chord-tangent law,
and $E_{ns} \cong K^*$ or $E_{ns} \cong K^+$ \cite{prop2.5silverman}.
Furthermore these isomorphisms are efficiently computable in either
direction.
(The latter case is not useful as discrete log is easy in $K^+$.)

For example, consider the curve
\[ E : y^2 = x^3 + x^2 \]
This has a singular point at $(0,0)$. Then the other solutions to
this curve (including the point at infinity) form a group
$E_{ns}(\Fqk)$ for any finite field $\Fqk$.

Furthermore, there is an isomorphism $E_{ns}(\Fqk) \rightarrow \Fqk^*$
given by
\[ (x,y) \mapsto \frac{y-x}{y+x} \]
(and maps $O$ to $1$)
which is efficienty computable in either direction.
A little algebra
shows the reverse map is
\[ z \mapsto \left( x', \frac{1+z}{1-z} x' \right) \]
where $x' = 4z/(1-z)^2$ for $z \ne 1$, and $1 \mapsto O$.

This example has little practical value as point additions are slower than
modular multiplications, and keeping both coordinates of a point in memory
takes twice as much space. However, it does suggest
that cryptography in finite fields can be viewed as a special
case of elliptic curve cryptography, and strictly speaking,
those that complain that elliptic curve cryptography is controversial,
too experimental and untested must qualify
their remarks by making exceptions for singular curves!

We note that if an efficiently bilinear nondegenerate pairing could be found
for singular curves then the Decisional Diffie-Hellman problem could be broken
in finite fields.

\section {Generating Random Points}

Let $E$ be an elliptic curve over a field $K$.
To generate a random point, first generate $x \in K$ randomly.
Next, we solve $E : Y^2 = X^3 + a X + b$ for $Y$.
Let $g(X) = x^3 + a X + b$. Then
we must find a square root of $g(x)$.

For a field of prime order one can use the Tonelli-Shanks algorithm
to compute square roots~\cite{bss, handbook},
which we quote here for handy reference.

Suppose we wish to compute $b = \sqrt{a}$ in
a field $\Fq$ for some prime $q$. (These paragraphs should be viewed
as self-contained; the notation from earlier paragraphs does not apply here.)

\begin{enumerate}
\item
Find an element $g \in \Fq$ that is not a square. This can be done
by choosing a random element $g$ until $g^{(q-1)/2} = -1$. 
Clearly this $g$ can be stored for use in later square root computations.
\item
Since $q$ is odd (unless $q = 2$ in which case square roots are trivial),
we may write $q - 1 = 2^s t$ for some odd $t$.
\item
$e \leftarrow 0$
\item
for $i \leftarrow 2$ to $s$ do
\begin{enumerate}
\item
if $(ag^{-e})^{(q-1)/2^i} \ne 1$ then $e \leftarrow 2^{i-1} + e$.
\end{enumerate}
\item
$h \leftarrow ag^{-e}$
\item
$b \leftarrow g^{e/2} h^{(t+1)/2}$
\item
Return $b$
\end{enumerate}

We can briefly explain the Tonelli-Shanks algorithm as follows.
Square roots in a cyclic group of order $t$ where $t$ is odd can be
computed by exponentiating by $(t+1)/2$.
The fact that $\Fq^* \cong \mathbb{Z}_{2^s}^+
\times \mathbb{Z}_t^+$ leads to the above.

For a general finite field,
one must use a more complex algorithm. Perhaps the simplest of these
Legendre's method which can be viewed as a special
case of the Cantor-Zassenhaus algorithm for factoring polynomials that
will be described in a later chapter.
Faster algorithms exist, though sometimes require precomputation~\cite{djb}.

Approximately half the time no square root exists.
In the case that $K$ has prime order,
one can compute the Legendre symbol before attempting to square root $g(x)$.
More generally it can be checked that $X^2 - g(x)$ is reducible. Alternatively,
one can proceed with a square root algorithm and square the result and
compare with $g(x)$.

In any event, if $g(x)$ is not a square then simply start again and
chose another random $x \in K$. Eventually a suitable $x$ will be found.
Once found, a coin is flipped that decides which of the square roots will
be the $y$-coordinate.

It is impossible to choose the point at infinity with this method,
which is usually desirable. Clearly, if one wishes to be able to pick
the point at infinity, one could simply add a step before choosing $x$:
with $1/n$ probability choose $O$ otherwise proceed normally, where
$n = \#E(K)$.

\section {Representing Points}

Instead of using a pair of numbers to represent a point,
we can use Jacobian projective coordinates: the triplet
$(x,y,z)$ represents the point $(x/z^2, y/z^3)$.

Then to double the point $(x,y,z)$, we compute $(x_3, y_3, z_3)$ by:

\[
\begin{array}{rcl}
\lambda_1 & \leftarrow & 3x^2 + a z^4 \\
z_3 & \leftarrow & 2yz \\
\lambda_2 & \leftarrow & 4xy^2 \\
x_3 & \leftarrow & \lambda_2^2 - 2 \lambda_2 \\
\lambda_3 & \leftarrow & 8 y^4 \\
y_3 & \leftarrow & \lambda_1(\lambda_2 - x_3) - \lambda_3
\end{array}
\]

We see the computation of $3x^2 + a z^4$ in general costs $3S + 1M$
where $M$ and $S$ denote the time required for a multiplication and
a squaring respectively. (We consider multiplication by
a small constant to be negligible.)

For $a=0$ the cost is only $1S$.

For small $a$ or fairly small $a$ with low Hamming weight the cost is $3S$.

For $a=-3$ we may compute $\lambda_1 = 3(x+z^2)(x-z^2)$ which costs
$1M + 1S$.

For $a = -3 d^2$ where $d$ is small or fairly small with low Hamming weight,
we may compute $\lambda_1 = 3(x+dz^2)(x-dz^2)$ which also costs $1M + 1S$.
By transforming the curve (see section ?!?)
we can obtain an $a$ that is faster than the general case.

TODO: quote them instead for handy reference:
We omit the formulas for projective point addition, and for adding
an affine point to a projective point, referring the reader instead
to Blake et al. or Cohen et al. \cite{stuff}.

The latter investigate other kinds of projective coordinates, where point
doubling or multiplication is even faster.

\section {Point Compression and Reduction}

We can optimize point representations for space instead of time,
useful for transmission and storage.
Given $x$, we can solve $Y^2 = x^3 + ax + b$ for $Y$, thus
a point $(x,y)$ can be represented by $x$ along with a single bit indicating
which solution of $y$ to take. This technique is known as \emph{point
compression}.
We discuss how to find square roots later.

Alternatively, for some cases we can simply use $x$ alone, which
is sometimes called \emph{point reduction}. For example,
suppose Alice needs to send Bob a point in some cryptosystem. She sends
Bob $x$, who guesses the solution $y$. Bob attempts to proceed normally.
If the protocol fails (e.g. the signature does not verify) then Bob tries
$-y$ and tries again. This does not cost much more since the solutions are
related by negation, and we will see that in some pairing-based systems
one can check both possibilities at the same time.

(Of course we can take this principle further. Alice can
omit $k$ bits of $x$, and leave Bob to try all $2^k$ possibilities.)

\section {Point Multiplication}

Point multiplication is the elliptic curve equivalent of modular
exponentiation, thus optimizations that apply to one often apply to the other.

One difference is that group inversion is much cheaper in elliptic curves:
one simply negates the $y$-coordinate, an operation that is essentially free,
whereas in finite fields one must spend time running the extended
Euclidean algorithm.

This allows one to use non-adjacent forms (or signed windows), which
TODO: refer to bss, handbook
Sliding Windows.

If it is known in advance that a particular point $P$ will feature in
several point mulitplications, preprocessing can be employed, namely
multiples of $P$ used often in the exponentiation routine
are computed and stored.

When computing expressions
of the form $a P + b Q$ (the equivalent of $g^x h^y$ in
finite fields), rather than computing $a P$ and $b Q$ separately
and adding the results, one can use
a multiexponentiation trick to roughly halve the number of point doublings
(if $a$ and $b$ are about the same length).
In its simplest form, $P + Q$ is precomputed and stored,
then a double-and-add algorithm is used to compute the
final result. Instead of describing the algorithm in full, we present
an example:

To compute $12 P + 7 Q$ one precomputes $P + Q$, then:

\begin{enumerate}
\item
$R \leftarrow P$
\item
$R \leftarrow 2R$
\item
$R \leftarrow R + (P + Q)$
\item
$R \leftarrow 2R$
\item
$R \leftarrow R + Q$
\item
$R \leftarrow 2R$
\item
$R \leftarrow R + Q$
\end{enumerate}

Multiexponentiation can
be generalized to computing any expression of the form
$a_1 P_1 + ... + a_t P_t$. Multiexponentiation can also be combined with
sliding window techniques.
One drawback is the amount of precomputation needed,
which worsens as $t$ increases, and also as the size of the sliding window
is increased in which case one must store $m P + n Q$ for several $m, n$.

Cite Lange Sica Quisquater

Cohen Miyaji Ono

\section {Security}

We have not yet shown any benefits to using elliptic curves instead
of finite fields. On the contrary, although a group inversion is
almost free, the group operation, hashing, and random element generation
are considerably more expensive for the same field.

Their strength derives from the fact that no discrete log algorithm for
general elliptic curves that outperforms generic methods has ever been
discovered.

The best generic methods are based on the birthday paradox, and have
$O(\sqrt{n})$ expected running time (where $n$ is the order of the group).
A group order around 160 bits in length is sufficient to defeat such attacks.
In contrast, subexponential discrete log algorithms for finite fields
mean that one must use at least 1024-bit finite fields for security.

Hence one may work with fields over six times smaller in length
if elliptic curves are used. The increase in speed in
the underlying field arithmetic easily compensates for the more
complex group operations. Thus choosing elliptic curves instead of
finite fields results in savings in both space and time.

The most important attribute of elliptic curves for our purposes
is that without them, we cannot construct a cryptographically
useful bilinear map.
However, we shall later find pairings require certain elliptic curves,
for which subexponential discrete log algorithms exist. In short,
one can map pairing-friendly elliptic curve groups to finite fields, where
index calculus can break the discrete log problem.

\section {Short Signatures}

At this point we can see why
the signature scheme introduced
in the first chapter features a short signature length.
Recall in abstract terms,
a signature is a single element of a cyclic group $G$ of prime order $r$
and we need a bilinear map to exist on $G$.

We shall see later
a concrete instantiation of such a group with a pairing
can be realized by using a subgroup of points
on an elliptic curve $E$ over some field $\Fq$ where
$q$ is about 160 bits long.

A point has two coordinates, both taking about 160 bits to represent.
We can use point reduction, that is, discard the $y$-coordinate entirely,
and represent signatures by their $x$-coordinate only if we modify
verification as follows:
\begin{enumerate}
\item
Given an $x$-coordinate of a signature, find any solution $y$
in the curve equation $E$.
\item
If the signature $(x,y)$ does not verify,
check if the signature $(x, -y)$ verifies.
\item
If the signature still does not verify then reject it.
\end{enumerate}

Thus signatures are elements of $\Fq$ and hence roughly 160 bits in length.

Verification of the signature involves computing $e(P, Q)$ for some points
$P, Q$. By the bilinearity of the pairing, if we have guessed the point $Q$
wrongly, we can obtain the value of $e(P,-Q) = e(P,Q)^{-1}$ by performing
an inversion rather than recompute another pairing.

Later we will see we can do better than this by using a technique known
as pairing compression or reduction, and in effect check both
cases at once.

Note instead of discarding the $y$-coordinate, we can
replace it with a single bit signifying which solution of $y$ to take.
