\beforepreface
\prefacesection{Preface}
Pairing-based cryptography is a relatively young area of cryptography that
revolves around a particular function with interesting propreties.
It allows the construction of novel cryptosystems that are otherwise
difficult or impossible to assemble using standard primitives.

We first define a cryptographic pairing abstractly and show how it
can be used to build a signature scheme that is simple yet has many
desirable properties.

Next we examine how a pairing can be implemented in practice. The only known
mathematical setting where suitable pairings exist are groups on certain
families of curves. We focus on the simplest and most well-understood case:
elliptic curves.

We discuss methods for finding curves that yield cryptographic pairings,
and outline various optimizations that can be applied. We shall see that
pairing-based cryptosystems are quite practical and compare well against
traditional cryptosystems.

It is hoped that this work will be useful guide to implementing pairing-based
cryptography to a programmer who has experience with conventional
cryptosystems. While not comprehensive (notably absent are in-depth
discussion of the characteristic 3 case, pairings where
the subgroup size is necessarily significantly smaller than the field size,
and hyperelliptic curves),
the information contained herein should be enough to allow one to build
practical pairing-based applications from scratch.

This text is intended to be self-contained. Aside from
arithmetic at the lowest level, algorithms not likely to
be found in a basic course in number theory or cryptography are quoted here,
enabling the reader to implement pairings without referring to any other
sources.

Some background is required by the reader as
we do not review basic abstract algebra.
On the other hand, we introduce enough elliptic curve theory for
cryptographic purposes.

\section {Summary of Contributions}

Our main original contributions are
\begin{itemize}
\item Abstract definitions of pairings that bridge the
gap between formal security proofs and bilinear maps used in practice.
\item A practical digital signature scheme with the shortest known
signature length at typical security levels. This signature scheme
in fact has many other desirable features, but they fall outside the scope
of this text.
\item The first published method for constructing cryptographically
useful pairings with any given embedding degree.
\item Optimizations that greatly improve the running time of a pairing
by at least a factor of four over a naive implementation. Some of our
techniques are now so commonplace that they are
sometimes considered "standard"~\cite{hsv}.
\end{itemize}

The following publications
form the foundation of this thesis:
\begin{itemize}
\item D.Boneh, B. Lynn, and H. Shacham. Short signatures from the Weil pairing.
{\it Journal of Cryptology}, 17(1):297--319, 2004.
\item
P. S. L. M. Barreto, H. Y. Kim, B. Lynn and M. Scott.
Efficient algorithms for pairing-based cryptosystems. In {\it CRYPTO 2002},
pages 354--368.
\item P. S. L. M. Barreto, B. Lynn and M. Scott.
Constructing elliptic curves with prescribed embedding degree. In {\it Third
Conference on Security in Communication Networks 2002}.
\item P. S. L. M. Barreto, B. Lynn and M. Scott.
On the selection of pairing-friendly groups. In {\it Selected Areas of Cryptography 2003}.
\end{itemize}
More polished versions of the above works can be found
in the Journal of Cryptology~\cite{blsjournal, balsjournal}.

Most of the described algorithms and optimizations
feature in the PBC (Pairing-Based Cryptography) library, maintained by the
author, and available under the GNU Public License at 
http://crypto.stanford.edu/ibe/
